Data Specialization for a Staged Lambda Calculus

### 2015-02-27 paper/splitting.tex
described next and prev splitting more technically

Consider any (except \texttt{prev}) 
$n$-ary term $e = \scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$.
First note that for any $\diatwo e q$, $q$ must have the form $\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$,
even under binders and for nullary terms.
For each $k \in \{1,\ldots,n\}$, the invariant for stage \bbtwo\ splitting of $e_k$ is,
\[
	\splittwo {e_k} A {p_k} {l_k} {r_k} \text{ and } \reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
\]
To match the invariant for a split $\splittwo {\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}} A p l r$, we require,
\[
	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
\]
Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
$l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.
Again, this works even for nullary terms and under binders.

Splitting \texttt{prev} generates a precomputation that projects the immediate
result of its world \bbone\ subterm.
Since the argument to \texttt{prev} is of $\fut$ type, its immediate result reduces to $\tup{}$, justifying why it can be thrown away.
Finally, splitting \texttt{next} simply tuples up the precomputation of its
world \bbtwo\ subterm with a trivial immediate result $\tup{}$.

>>>>

Consider any (except \texttt{prev}) 
$n$-ary term $e = \scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}$.
First note that for any $\diatwo e q$, $q$ must have the form $\scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}$,
even under binders and for nullary terms.
For each $k \in \{1,\ldots,n\}$, the invariant for stage \bbtwo\ splitting of $e_k$ is,
\[
	\splittwo {e_k} A {p_k} {l_k} {r_k} \text{ and } \reduce {p_k} {b_k} \text{ and } [{b_k}/{l_k}]{r_k} \equiv {q_k}
\]
To match the invariant for a split $\splittwo {\scriptCapp {e_1 \ttsemi \ldots \ttsemi e_n}} A p l r$, we require,
\[
	\reduce p b \text{ and } [b/l]r \equiv \scriptCapp {q_1 \ttsemi \ldots \ttsemi q_n}
\]
Our algorithm accomplishes this by using $p = \tup {p_1, \ldots, p_n}$ and
$l.r = \tup {l_1, \ldots, l_n}.\scriptCapp {r_1 \ttsemi \ldots \ttsemi r_n}$.

Again, this holds even for nullary terms and under binders.
In particular, \texttt{case} splits according to the rule:
\[
\infersplittwo	{	\begin{array} {l}
						\caseof{e_1}{x_2.e_2}{x_3.e_3} \splittwosym \\
						\hspace{5em} \pipeS {\tup{p_1,p_2,p_3}} {\tup{l_1,l_2,l_3}} {\caseof{r_1}{x_2.r_2}{x_3.r_3}}		
					\end{array}}	
				{\sub 1 ? & \sub 2 ? & \sub 3 ?} 								
\]
Note how the precomputations are pulled out from underneath the binder into a tuple that is always executed in the first stage,
regardless of what branch is eventually executed in the second stage.

\paragraph{Next and Prev}
Since \texttt{next} and \texttt{prev} do nothing more than convert between 
second stage terms at \bbtwo\ and second stage terms encapsulated within $\fut$s at \bbonem,
their splitting rules are relatively simple.

Given an $\splittwo e ? p l r$ and $\diatwo e q$ for which the splitting invariant holds, we know
\[
	\reduce p b \text{ and } [b/l]r \equiv q
\]
The splitting invariant for $\splitone {\next {e'}} ? {c'} {l'.r'}$ 
and $\diaone {\next {e'}} {\yhat \mapsto q} {\next \yhat}$
and $\rtab {\yhat \mapsto q} {\next \yhat} \vsplito \mval {\tup{}} {\letin \yhat q \yhat}$
is
\[
	\reduce {c'} {\tup {\tup{},b'}} \text{ and } [b'/l']r' \equiv \letin \yhat q \yhat
\]
which can be fulfilled using $c' = \tup{\tup{},p}$ and $l.r = l'.r'$.
Applying this process in reverse yields a splitting rule for $\prev e$ which applies \texttt{pi1} to the combined term of $e$.
